A formal approach but practical approach in train regulation

### 2025-06-20 paper/main.tex
Ajout règles du train

\documentclass[runningheads]{llncs}
%
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{wrapfig}
\usepackage{package/mathpartir}
\usepackage{todonotes}
\usepackage{xspace}

@@

\usepackage{package/formal-grammar}
\usepackage{package/mathpartir}

@@

% Je suis assez d'avis de distinguer la direction des blocks et des trains
\newcommand{\forward}{\ensuremath{\mathtt{f}}}
\newcommand{\backward}{\ensuremath{\mathtt{b}}\xspace}
\newcommand{\sucblock}{{\mathtt{suc}}}
\newcommand{\networkConf}{\ensuremath{\Sigma}}

%% Train directions
\definecolor{dircolor}{rgb}{1.0, 0.13, 0.32}
\newcommand{\dirFmt}[1]{{\color{dircolor}{#1}}}
\newcommand{\dirForward}{\ensuremath{\dirFmt{\forward}}\xspace}
\newcommand{\dirBackward}{\ensuremath{\dirFmt{\backward}}\xspace}
\newcommand{\dirStop}{\ensuremath{\dirFmt{\star}}\xspace}

%% Positions
\definecolor{poscolor}{rgb}{0.6, 0.45, 0.32}
\newcommand{\posFmt}[1]{{\color{poscolor}{#1}}}
\newcommand{\blocks}{{\posFmt{\mathsf{B}}}}
\newcommand{\bid}[1]{\ensuremath{\posFmt{b_{#1}}}}

%% Switches
\definecolor{swcolor}{rgb}{0.4, 0.23, 0.56}
\newcommand{\swFmt}[1]{{\color{swcolor}{#1}}}
\newcommand{\sid}[1]{\ensuremath{\swFmt{s_{#1}}}}
\newcommand{\turnouts}{\swFmt{\mathsf{\theta}}}
\newcommand{\switches}{\ensuremath{\swFmt{\sigma}}}

%% Trains
\definecolor{traincolor}{rgb}{0.2, 0.2, 0.6}
\newcommand{\trainFmt}[1]{{\color{traincolor} #1}}
\newcommand{\trainTuple}[4]{\langle \trainFmt{#1}, \posFmt{#2}, \dirFmt{#3}, #4 \rangle}
\newcommand{\trainSeq}{\trainFmt{\Gamma}}
\newcommand{\trains}{{\trainFmt{\mathsf{T}}}}
\newcommand{\tid}[1]{\ensuremath{\trainFmt{t_{#1}}}}

@@

%% Guard
\definecolor{guardcolor}{rgb}{0.53, 0.66, 0.42}
\newcommand{\guardFmt}[1]{{\color{guardcolor} \ensuremath{\mathtt{#1}}}}
\newcommand{\guardT}{\guardFmt{T}}
\newcommand{\guardS}{\guardFmt{S1}}
\newcommand{\guardSS}{\guardFmt{S2}}
\newcommand{\guardR}{\guardFmt{R}}

@@

%% Reduction relation
\newcommand{\reduces}{\ensuremath{\rightarrow}}

@@

\newcommand{\su}[2]{{\mbox{$\mathtt{startuntil}(\dirFmt{#1}, \posFmt{#2})$}}\xspace}

@@

The network is then described as a triple $(\blocks, \turnouts, \sucblock)$ where $\sucblock$ is a partial function indicating the following block given a starting position, a direction and a network configuration: $\sucblock: \blocks\times \directions \times \networkConf \to \blocks$. We note $\sucblock(pid, d, \sigma) = \bot$ when it is undefined. From an abstract point of view, the topology of the network reduces to that data.

@@

Essentially, this guard is the state machine shown in Figure~\ref{fig:state_machine_guard}. The four states of the guard are \guardR, \guardS, \guardSS, and \guardT.

\begin{figure}
	\centering
	\begin{tikzpicture}
		\node[draw] (trains) {Train movements (\guardT)};
		\node[draw, above right=of trains] (signals1) {Signals update (\guardS)};
		\node[draw, below right=of trains] (regulator) {Regulator reaction (\guardT)};
		\node[draw, right=of regulator] (signals2) {Signals update (\guardSS)};

		\draw[-stealth] (trains) to[bend right=20] (signals1);
		\draw[-stealth] (signals1.west) to[bend right=20] (trains.north);
		\draw[-stealth] (trains.south) to[bend right] (regulator.west);
		\draw[-stealth] (regulator) to (signals2);
		\draw[-stealth] (signals2) to[bend right=10] (trains);
	\end{tikzpicture}
	\caption{State machine of the guard.}
	\label{fig:state_machine_guard}
\end{figure}

@@

>>>>

\documentclass[runningheads]{llncs}
%
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{wrapfig}
\usepackage{package/mathpartir}
\usepackage{todonotes}
\usepackage{xspace}

@@

\usepackage{package/formal-grammar}

\newcommand{\ruleFmt}[1]{\textsc{(#1)}}
\newcommand{\ruleDef}[1]{\hypertarget{#1}{\ruleFmt{#1}}}
\newcommand{\ruleRef}[1]{\hyperlink{#1}{\ruleFmt{#1}}}

@@

% Je suis assez d'avis de distinguer la direction des blocks et des trains
\newcommand{\forward}{\ensuremath{\mathtt{f}}}
\newcommand{\backward}{\ensuremath{\mathtt{b}}\xspace}
\newcommand{\sucblock}{{\mathtt{suc}}}

%% Train directions
\definecolor{dircolor}{rgb}{1.0, 0.13, 0.32}
\newcommand{\dirFmt}[1]{{\color{dircolor} #1}}
\newcommand{\dirForward}{\ensuremath{\dirFmt{\forward}}\xspace}
\newcommand{\dirBackward}{\ensuremath{\dirFmt{\backward}}\xspace}
\newcommand{\dirStop}{\ensuremath{\dirFmt{\star}}\xspace}

%% Positions
\definecolor{poscolor}{rgb}{0.6, 0.45, 0.32}
\newcommand{\posFmt}[1]{{\color{poscolor}{#1}}}
\newcommand{\blocks}{{\posFmt{\mathsf{B}}}}
\newcommand{\bid}[1]{\ensuremath{\posFmt{b_{#1}}}}
\newcommand{\suc}[3]{\ensuremath{\sucblock(\posFmt{#1}, \dirFmt{#2}, \swFmt{#3})}}

%% Switches
\definecolor{swcolor}{rgb}{0.4, 0.23, 0.56}
\newcommand{\swFmt}[1]{{\color{swcolor}{#1}}}
\newcommand{\sid}[1]{\ensuremath{\swFmt{s_{#1}}}}
\newcommand{\turnouts}{\swFmt{\mathsf{\theta}}}
\newcommand{\switches}{\ensuremath{\swFmt{\sigma}}}
\newcommand{\networkConf}{\ensuremath{\swFmt{\Sigma}}}
\newcommand{\nosuc}{\ensuremath{\swFmt{\bot}}}

%% Trains
\definecolor{traincolor}{rgb}{0.2, 0.2, 0.6}
\newcommand{\trainFmt}[1]{{\color{traincolor} #1}}
\newcommand{\trainTuple}[4]{\langle \trainFmt{#1}, \posFmt{#2}, \dirFmt{#3}, #4 \rangle}
\newcommand{\trainSeq}{\trainFmt{\Gamma\xspace}}
\newcommand{\trains}{{\trainFmt{\mathsf{T}}}}
\newcommand{\tid}[1]{\ensuremath{\trainFmt{t_{#1}}}}

%% Train program
\newcommand{\su}[2]{{\mbox{$\mathtt{startuntil}(\dirFmt{#1}, \posFmt{#2})$}}\xspace}
\newcommand{\trainConcat}[2]{#1 \cdot #2}
\newcommand{\emptyTrainProg}{\varepsilon}

@@

%% State
\newcommand{\stateTuple}[4]{\tuple{#1, #2, #3, #4}}

%% Guard
\definecolor{guardcolor}{rgb}{0.53, 0.66, 0.42}
\newcommand{\guardFmt}[1]{{\color{guardcolor} \ensuremath{\mathtt{#1}}}}
\newcommand{\guardT}{\guardFmt{T}}
\newcommand{\guardS}{\guardFmt{S}}
\newcommand{\guardR}{\guardFmt{R}}

@@

%% Reduction relation
\newcommand{\reduces}{\ensuremath{\rightarrow}}
\newcommand{\redTuple}[3]{\ensuremath{\tuple{#1, #2, #3}}}

@@

@@

The network is then described as a triple $(\blocks, \turnouts, \sucblock)$ where $\sucblock$ is a partial function indicating the following block given a starting position, a direction and a network configuration: $\sucblock: \blocks\times \directions \times \networkConf \to \blocks$. We note $\suc{\bid{id}}{d}{\switches} = \nosuc$ when it is undefined. From an abstract point of view, the topology of the network reduces to that data.

@@

Essentially, this guard is the state machine shown in Figure~\ref{fig:state_machine_guard}. The three states of the guard are \guardR, \guardS, and \guardT.

\begin{figure}
	\centering
	\begin{tikzpicture}
		\node[draw] (trains) {Train movements (\guardT)};
		\node[draw, above right=of trains] (signals1) {Signals update (\guardS)};
		\node[draw, below right=of trains] (regulator) {Regulator reaction (\guardR)};

		\draw[-stealth] (trains) to[bend right=20] (signals1);
		\draw[-stealth] (signals1.west) to[bend right=20] (trains.north);
		\draw[-stealth] (trains.south) to[bend right] (regulator.west);
		\draw[-stealth] (regulator) to[bend right] (signals1);
	\end{tikzpicture}
	\caption{State machine of the guard.}
	\label{fig:state_machine_guard}
\end{figure}

@@

\paragraph{Train rules.}
We first focus on rules that affect trains. The first rule is \ruleRef{Start}, which is used to change the direction of a train (e.g. to go from a stop to a given direction, or to change direction if the train has to backtrack). This rule can be triggered when the program of the train begins by \su{d^\prime}{N}. This order says that the train should got in direction $\dirFmt{d^\prime}$, and on that direction, it is expected to meet the sequence of block which ids are in $\posFmt{N}$.

\begin{mathpar}
	\inferrule*[Left=\ruleDef{Start}]{
		\dirFmt{d} \neq \dirFmt{d^\prime}
	}{
		\redTuple{\guardT}{\bufferFmt{B}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{d}{\trainConcat{\su{d^\prime}{\posFmt{N}}}{P^\prime}}\}}{\regulator}{\switches}{\signals}} \\
		\reduces
		\redTuple{\guardT}{\bufferFmt{B}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{d^\prime}{\trainConcat{\su{d^\prime}{\posFmt{N}}}{P^\prime}}\}}{\regulator}{\switches}{\signals}}
	}
\end{mathpar}

Conversely, if the train program is empty (i.e. upon termination), the train stops using rule \ruleRef{Stop}.
\begin{mathpar}
	\inferrule*[Left=\ruleDef{Stop}]{
		\dirFmt{d} \neq \dirStop
	}{
	\redTuple{\guardT}{\bufferFmt{B}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{d}{\emptyTrainProg}\}}{\regulator}{\switches}{\signals}} \\
	\reduces
	\redTuple{\guardT}{\bufferFmt{B}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{\dirStop}{\emptyTrainProg}\}}{\regulator}{\switches}{\signals}}
	}
\end{mathpar}

The last two rules for trains describe what happens when a train actually moves, i.e. when its position changes. Two cases can occur, depending on the current order of the train: in \su{d}{N}, the $\posFmt{N}$ parameter is a list of blocks the train is about to encounter (including the current block), while going in direction \dirFmt{d}. Two cases occur, depending on whether \posFmt{N} contains future blocks or not. If \posFmt{N} contains only the upcoming block, then upon leaving the block, the order is removed (rule \ruleRef{UntilNext}); otherwise the upcoming block is simply popped from \posFmt{N} (rule \ruleRef{Until}). Notice that in both cases, the state of the guard changes to \guardS, meaning that signals will be updated immediately after that rule.\todo{Introduire/formaliser N}

\begin{mathpar}
	\inferrule*[Left=\ruleDef{Until}] {
		\suc{\bid{id}}{d}{\switches} = \bid{id^\prime} \neq \nosuc
		\and
		\signalF{\bid{id}}{d} = \siggreen
		\and
		\posFmt{N}\neq \varepsilon
	}{
		\redTuple{\guardT}{\bufferFmt{B}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{d}{\trainConcat{\su{d}{[\bid{id^\prime}; \posFmt{N}]}}{P}}\}}{\regulator}{\switches}{\signals}} \\
		\reduces		
		\redTuple{\guardS}{\bufferFmt{B}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id^\prime}}{d}{\trainConcat{\su{d}{[\posFmt{N}]}}{P}}\}}{\regulator}{\switches}{\signals}}
	}
\end{mathpar}

\begin{mathpar}
	\inferrule*[Left=\ruleDef{UntilNext}] {
		\suc{\bid{id}}{d}{\switches} = \bid{id^\prime} \neq \nosuc
		\and
		\signalF{\bid{id}}{d} = \siggreen
	}{
		\redTuple{\guardT}{\bufferFmt{B}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{d}{\trainConcat{\su{d}{[\bid{id^\prime}]}}{P}}\}}{\regulator}{\switches}{\signals}} \\
		\reduces		
		\redTuple{\guardS}{\bufferFmt{B}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id^\prime}}{d}{P}\}}{\regulator}{\switches}{\signals}}
	}
\end{mathpar}

\paragraph{Regulator rules.}
